---
title: "Week6-Project2"
author: "Musrat Jahan"
date: "`r Sys.Date()`"
output: 
    html_document:
    theme: cerulean
    toc: true
    toc_float: true
    number_sections: false
    toc_collapsed: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(viridis)
library(RSocrata)
library(readr)
library(readxl)
library(janitor)
library(lubridate)
```

# Section 1: Emissions Data 

## Load data set
```{r}
df <- read.csv("https://raw.githubusercontent.com/justin-2028/Total-Emissions-Per-Country-2000-2020/refs/heads/main/Total%20Emissions%20Per%20Country%20(2000-2020).csv")


colnames(df) <- gsub("^X", "", colnames(df))

head(df)
```

## Make longer
I changed all the year columns to one column under year. I made the data set longer. This makes it tidy. 
```{r}
df_longer <- df |> 
  pivot_longer(
    cols = starts_with("2"), 
    names_to = "year", 
    values_to = "total emissions",
    values_drop_na = TRUE
  )


```

## Total emissions per country for each year 
```{r}
yearly_emissions_by_area <- aggregate(df_longer$'total emissions', by = list(df_longer$year, df_longer$Area), FUN = sum)

yearly_emissions_by_area

#rename columns
yearly_emissions_by_area <-
yearly_emissions_by_area %>% 
  rename(
    year = Group.1,
    country = Group.2,
    emissions = x
    )

yearly_emissions_by_area
```

## Analyze overall total emissions per country for each year 
Too many different countries 
```{r}
ggplot(yearly_emissions_by_area, aes(x = year, y = emissions, 
                                     fill = country)) +
  geom_tile()
```


## Analyze total emissions over time
As you can see from the graph, total emissions have gone up steadily from 2000 to 2019, but in 2020, it decreased a significant amount. This might be due to more awareness about climate change and global warming.
```{r}
yearly_emissions <- aggregate(df_longer$'total emissions', by=list(df_longer$year), FUN = sum)
yearly_emissions

#rename columns
yearly_emissions <-
yearly_emissions %>% 
  rename(
    year = Group.1,
    emissions = x
    )

ggplot(yearly_emissions, aes(x = year, y = emissions)) +
  geom_point()
```

## Total emissions per country 

Some of the top countries that contributed to emissions are China,  USA, Brazil, India, Indonesia, and Democratic Republic of the Congo.  
```{r}
emissions_by_area <- aggregate(df_longer$'total emissions', by = list(df_longer$Area), FUN = sum)


#rename columns
emissions_by_area <-
emissions_by_area %>% 
  rename(
    country = Group.1,
    emissions = x
    )

top <- emissions_by_area[order(-emissions_by_area$emissions),]
head(top, 50)
```

## Analysis of Total emissions per country 
too many countries, cant read 
```{r}
ggplot(emissions_by_area, aes(x = country, y = emissions), label=NA)+ geom_point() 

```


## Aanalyze by emission type
Emissions (CO2eq) (AR5) are highest. They are over 200,000 kilotonnes. The second highest place is tied with emissions (N20) and emissions (CO2eq) from N20 (AR5). Lowest emissions are (CO2eq) from F-gases, less than 25,000 kilotonnes.
```{r}

ggplot(df_longer, aes(y=Element)) +
         geom_bar()
```

## USA emission types distribution
The distribution looks very similar to the distribution with the data from all the regions. For USA, the counts are smaller. Highest are emissions (CO2eq) (AR5), a little less than 1250 kilotonnes. Lowest emissions are (CO2eq) from F-gases, around 125 kilotonnes. The second highest place is from emissions (N20), around 950 kilotonnes.
```{r}
usa <- df_longer %>%
  filter(Area == "United States of America")

ggplot(usa, aes(y=Element)) +
         geom_bar()
```

## Item Analysis
Highest item is IPCC Agriculture. Second highest is farm gate emissions.
Lowest is international bunkers.

```{r}

ggplot(df_longer, aes(y=Item)) +
         geom_bar()
```

## Item Analysis USA
Highest item is IPCC Agriculture, just like in the overall data. Lowest is drained organic soils (C02) and drained organic soils.

```{r}
ggplot(usa, aes(y=Item)) +
         geom_bar()
```

# Section 2: NPDES NY Data

### Loading NYSDEC DART Data
[insert info about data here]
```{r call-dart-api}
# API call to NYS DEC DART
# would be nice to call the API for NPDES permits (as listed in p_type)
# dart <- read.socrata("https://data.ny.gov/resource/mbk7-f2r2.json")
# write_csv(dart,"dart_3.7.2025.csv")

# note: this csv is too large to upload - will create a new csv.
dart <- read_csv("dart_3.7.2025.csv")
```

### Defining Regex Patterns to Detect NPDES IDs
[insert info here about IDs]
```{r defining-regex-patterns}
# p_type = permit type, in this exercise we are filtering for wastewater permits
p_type <- c("P/C/I SPDES - Surface Discharge",
            "Municipal SPDES - Surface Discharge",
            "Industrial SPDES - Surface Discharge")

# defining the regex patterns for the IDs we want to track
npdes_pattern <- "NY\\d{7}"
gp_pattern <- "GP\\d{7}"
individual_pattern <- "NY[A-Z]\\d{2}[A-Z]\\d{3}"
all_patterns <- paste(npdes_pattern,gp_pattern,individual_pattern, sep="|")
```

### Creating the NPDES Universe 
```{r creating-universe-v1}
# creating the permit universe pulling from NYSDEC's DART System 
# and detecting the string patterns within DART to assign permit type: npdes, 
# individual(i.e., a permit covered under a general permit), general, 
# or multi (meaning the DART entry had multiple associated IDs).

universe <- dart |>  
  filter(`permit_type` %in% p_type) |>  
  mutate(
    npdes = str_count(`other_known_ids`, npdes_pattern),
    individual = str_count(`other_known_ids`, individual_pattern),
    gp =  str_count(`other_known_ids`, gp_pattern),
    sum_ids = rowSums(across(c(`npdes`, `individual`,`gp`))),
    npdes_id = str_extract_all(`other_known_ids`, all_patterns),
    date_received=as.Date(date_received, format = "%d-%m-%Y")
    ) |> 
  mutate(applicant_id =cur_group_id(),.by = applicant) |> # creating applicant id
  mutate(facility_id = cur_group_id(),.by = c(facility,location,town_or_city)) |> # creating facility id
  distinct() |> # removing duplicate rows 
  mutate(
          dart_p_type = case_when(sum_ids  > 1 ~ "multi",
                               sum_ids & npdes == 1 ~ "npdes",
                               sum_ids & individual == 1 ~ "individual",
                               sum_ids & gp == 1 ~ "gp")) |>  
  unnest_longer(npdes_id, keep_empty = FALSE) |> 
  filter(!is.na(npdes_id))

```
Note: The code above filters entries that did not have a NPDES ID listed in the 
"Other Known IDs" column, however, were listed as NPDES permits in the Permit 
Type Column. However, out of 35,642 entries, only 69 were missing NPDES IDs. 

```{r note-on-str-count}
# the str_count statements are taking a sum of associated permits from DART data 
# entries. If a permit is associated with multiple ids it is assigned multi, 
# if it is a regular npdes permit, it is called npdes. If the permit is an 
# individual permit under a general permit, it is called individual. if it is a 
# general permit, it is called 'gp'. for information about permit types please 
# see the "What is the process for applying for coverage under a NPDES general 
# permit" section of the EPA NPDES Permit Basics Page.
# https://www.epa.gov/npdes/npdes-permit-basics
```

### Table 1: Permit Level Data
```{r tbl1-permit-lvl}
# this table shows the most recent permit information
tbl1_permit_lvl <- universe  |> 
  group_by(npdes_id) |> 
  slice(which.max(date_received)) |>
  select(npdes_id,facility_id,application_id,applicant,applicant_id,permit_type,
         status,date_received,upa_class,seqr_class,seqr_determination,
         lead_agency,coastal_zone_status, final_disposition,permit_effective_date,
         permit_expration_date,dec_contact,shpa_status,enivronmental_justice)
```

### Table 2: Permit Action Level Data 
```{r tbl2-permit-action-lvl}
# this table shows the permit history. each observation in this table represents
# a permit action.

# if there are true duplicates (meaning all rows are equal, then the duplicate
# row can be dropped)

tbl2_permit_act_lvl <- universe |> 
    mutate(action_id = paste(npdes_id,date_received, sep = "_")) |> 
    distinct() |> 
    mutate(dup_flag = duplicated(action_id),
           transfer_flag=str_detect(toupper(short_description),"TRANSFER")) |> 
  select(action_id,facility,facility_id,npdes_id,application_id,applicant,
         application_type,date_received,status,short_description,
         enb_publication_date,written_comments_due,dup_flag,transfer_flag)
```

### Table 3: Facility Level Data
```{r tbl3-facility-table}
# this table shows the facility information. Each observation in this table
# represents a facility associated with NPDES permits.
tbl3_facility_lvl <- universe |> 
  select(facility_id,facility,
         location,town_or_city) |> 
  distinct() |> 
  arrange(facility_id)
```

### Table 4: NPDES Permit Applicant Table 
```{r tbl4-applicant-table}
# this table shows the applicant information. Each observation in this table
# represents a permit applicant for NPDES permits.
tbl4_app_lvl <- universe |> 
  group_by(applicant_id) |> 
  slice(which.max(date_received)) |>
  select(applicant_id,applicant,application_id)
```

## Data Tables and Structure

        (1) tbl1 - permit table: the purpose of this table is to have the most
        recent permit information. This will have one row per permit.
        
        (2) tbl2 - permit action table: the purpose of this table is to have a 
        table with every permit-action. This means there should be one row per
        permit action. 
        
        (3) tbl3 - facility table: the purpose of this table is to have 
        information on the facility. 
        
        (4) tbl4 - applicant table: the purpose of this table is to have 
        information about the applicant.  

## Data Considerations:
- There was missing data, such as NPDES IDs. This means that some permit 
information may not be available. 

- There may be facilities that are listed as different facilities due to address
changes. This information should be paired with other facility information, like
EPA's Enforcement and Compliance History Online 
(ECHO, available at: https://echo.epa.gov/).

- For entries that were made on the same day for a particular permit, it is not 
possible to identify which entry was made first. Permit transfer actions are 
largely affected by this. Due to this, duplicates and transfers are flagged for 
manual review. 

Data Analysis Questions to Answer:
- Which facilities have multiple NPDES permits associated with it?
- Which applicants are associated with multiple NPDES permits?
- How do we identify if a facility has numerous permits? 

