---
title: "Project 2: Data Transformation"
author: "Group Members: Alina Vikhnevich, Olivia Azevedo, Alyssa Gurkas, Musrat Jahan"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
```

# **Introduction**

*TODO: Provide an introduction explaining the objective of the project, an overview of the datasets, and the transformation goals.*

---

# **Dataset 1: Placeholder Name**

## **Load Data**
```{r load_data1}
# TODO: Read in Dataset 1 (Paste dataset here)
dataset1 <- read_csv("path/to/dataset1.csv")
```

## **Initial Exploration**
```{r explore_data1}
# TODO: Perform initial exploration
glimpse(dataset1)
summarise_all(dataset1, ~sum(is.na(.)))
```

## **Data Cleaning & Transformation**
```{r transform_data1}
# TODO: Apply tidying steps using dplyr and tidyr
# dataset1 <- dataset1 %>% ...
```

## **Analysis**
```{r analyze_data1}
# TODO: Perform required analysis for Dataset 1
```

---

# **Dataset 2: Placeholder Name**

## **Load Data**
```{r load_data2}
# TODO: Read in Dataset 2 (Paste dataset here)
dataset2 <- read_csv("path/to/dataset2.csv")
```

## **Initial Exploration**
```{r explore_data2}
# TODO: Perform initial exploration
glimpse(dataset2)
summarise_all(dataset2, ~sum(is.na(.)))
```

## **Data Cleaning & Transformation**
```{r transform_data2}
# TODO: Apply tidying steps using dplyr and tidyr
# dataset2 <- dataset2 %>% ...
```

## **Analysis**
```{r analyze_data2}
# TODO: Perform required analysis for Dataset 2
```

---

# **Dataset 3: Placeholder Name**

## **Load Data**
```{r load_data3}
# TODO: Read in Dataset 3 (Paste dataset here)
dataset3 <- read_csv("path/to/dataset3.csv")
```

## **Initial Exploration**
```{r explore_data3}
# TODO: Perform initial exploration
glimpse(dataset3)
summarise_all(dataset3, ~sum(is.na(.)))
```

## **Data Cleaning & Transformation**
```{r transform_data3}
# TODO: Apply tidying steps using dplyr and tidyr
# dataset3 <- dataset3 %>% ...
```

## **Analysis**
```{r analyze_data3}
# TODO: Perform required analysis for Dataset 3
```

---

# **Final Analysis & Discussion**
```{r final_analysis}
# TODO: Compare results across datasets and summarize key findings
```

# **Conclusion**

*TODO: Summarize key takeaways, challenges encountered, and the importance of tidying data.*
