---
title: "Week 4 / Assignment – Tidying and Transforming Data"
author: "Alina Vikhnevich"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup_main, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
library(dplyr)
library(tidyr)
library(readr)
```

## **Introduction**
This report focuses on tidying and transforming data in R. Using `tidyr` and `dplyr`, we'll reshape data between wide and long formats, making it more suitable for analysis. 

## **Loading Data**
```{r load_data}
flight_data <- read_csv("/Users/alina_vikhnevich/Desktop/Spring 2025/DATA 607/DATA607/flight_data.csv")
print(flight_data)
```

## **Tidying the Data**
### **Reshaping Wide to Long Format**
```{r wide_to_long}
tidy_flight_data <- flight_data %>% 
  pivot_longer(cols = -c(Airline, Status), 
               names_to = "Destination", 
               values_to = "Count")
print(tidy_flight_data)
```

## **Summarizing the Data**
### **Total Flights by Airline and Status**
```{r summarize_data}
summarized_data <- tidy_flight_data %>% 
  group_by(Airline, Status) %>% 
  summarise(Total_Flights = sum(Count), .groups = 'drop')
print(summarized_data)
```

## **Visualizing the Data**
### **Bar Plot: On-Time vs Delayed Flights**
```{r plot_flight_status, echo=FALSE}
library(ggplot2)
ggplot(summarized_data, aes(x=Airline, y=Total_Flights, fill=Status)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="On-Time vs Delayed Flights by Airline", x="Airline", y="Total Flights") +
  theme_minimal()
```

## **Conclusion**
This assignment demonstrated how to reshape and analyze flight data in R. Using `tidyr` and `dplyr`, we converted wide-format data into a tidy structure and summarized key insights. This process makes data easier to work with for visualization and analysis.

