---
title: "DATA 607 Final Project: Pop Culture & Public Search Behavior"
author: "Group Members: Alina Vikhnevich, Olivia Azevedo, Alyssa Gurkas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    toc_collapsed: true
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Packages

```{r setup, message=FALSE, warning=FALSE}
library(bigrquery)
library(DBI)
library(tidyverse)
library(tidytext)
library(lubridate)
library(wordcloud)
library(topicmodels)
library(countrycode)
library(textmineR)
library(stm)
```

This project uses several R packages to support data acquisition,
cleaning, analysis, and visualization. For example, `bigrquery` and
`DBI` are used to query data from Google BigQuery, `tidytext` enables
text mining and topic modeling, and `ggplot2` is used for data
visualization.

# Introduction

In this project, we explore how search behavior in the United States
reflects interest in major pop culture events. Using unsupervised topic
modeling, we analyze Google’s top search terms to identify recurring
themes and trends. By linking search activity with globally documented
events, we aim to uncover how societal attention manifests in digital
behavior.

## Data Sources

-   **Google BigQuery – Top 25 Search Terms**\
    Google’s public dataset on the top 25 search queries per day,
    available via BigQuery.
-   **GDELT 2.0 – Our Global World in Realtime**\
    A large-scale global event database tracking news events from around
    the world in real time, including media coverage of pop culture,
    conflict, disasters, and more.

## Data Acquisition

### 1. Google BigQuery – Top 25 Search Terms

We use the `bigrquery` package to access Google’s public dataset of the
most popular daily search terms. Make sure you have a Google Cloud
project set up with billing enabled and a BigQuery API key.

```{r bigquery-auth, eval=FALSE}
# Authenticate with Google Cloud (interactive)
bq_auth(path = "data607finalproject-6c666e5b6214.json")
```

```{r google-bigquery-api}
project_id <- "data607finalproject"

# gbigquery is in GMT so if we do sub date -1 that means that it is taking the 
# data from today, need to do -2
# need to specify that the date is in America/NY time: CURRENT_DATE(‘America/New_York’)

# dma = designated market area
# need to pick two dmas if want to use score **
query <- as.character(
  "SELECT
  'top_terms' as term_type,
  term,
  score,
  refresh_date,
  dma_id,
  dma_name,
  ARRAY_AGG(STRUCT(rank,week) ORDER BY week DESC LIMIT 1) x
FROM
    `bigquery-public-data.google_trends.top_terms`
WHERE refresh_date >= date_add(current_date('America/New_York'),interval - 2 week)
AND regexp_contains(dma_name, 'New York|Los Angeles')
GROUP BY ALL

union  all

SELECT
  'rising_terms' as term_type,
  term as term_rising,
  score as score_rising,
  refresh_date,
  dma_id,
  dma_name,
  ARRAY_AGG(STRUCT(rank,week) ORDER BY week DESC LIMIT 1) x
FROM
    `bigquery-public-data.google_trends.top_rising_terms`
WHERE refresh_date >= date_add(current_date('America/New_York'),interval - 2 week)
AND regexp_contains(dma_name, 'New York|Los Angeles')
GROUP BY ALL
ORDER BY
    (select week from unnest(x)) desc,
    (SELECT rank FROM UNNEST(x)),
    refresh_date desc
"
)

# Run dry run query to estimate billing
job <- bq_perform_query_dry_run(
  billing = project_id,
  query = query,
  use_legacy_sql = FALSE
)

google_search_data <- bq_project_query(project_id, query)
google_search_df <- bq_table_download(google_search_data)
unnest_google_search_df <- google_search_df  |> 
  unnest(x)

head(unnest_google_search_df)
```

### 2. GDELT 2.0: Our Global World in Realtime

**Getting the latest file name:**

```{r gdelt-latest-link}
# Get the name of the most recent GDELT CSV file
latest_file <- read_lines("http://data.gdeltproject.org/gdeltv2/lastupdate.txt")
latest_file
```

**Download and Unzip:**

```{r gdelt-download}
# Constructing the full URL to download the ZIP file
export_url_line <- latest_file[1]

# Extract the URL from the line using strsplit
export_url <- strsplit(export_url_line, " ")[[1]][3]
export_url

# Download the export ZIP file to your working directory
download.file(export_url, destfile = "Data/GDELT/gdelt_latest.zip", mode = "wb")

# Unzip the file
unzip("Data/GDELT/gdelt_latest.zip", exdir='Data/GDELT')
```

**Read CSV into R:**

```{r gdelt-extract}
# Extract filename only from URL
csv_file <- paste0('Data/GDELT/', gsub(".zip", "", basename(export_url)))


# Read the CSV file with correct case-sensitive name
gdelt_df <- read_csv(csv_file, col_names = FALSE)

# Preview data
head(gdelt_df)
```

**GDELT 2.0 – Structured Load**

GDELT 2.0 is a massive global event database that captures media
coverage of events in real time from across the world. The export files
are updated every 15 minutes and are tab-delimited without column
headers, so we must assign names manually using the GDELT schema.

In this chunk, we load the most recent export file and apply column
names for easier exploration:

```{r gdelt-read}
# Column names based on GDELT 2.0 schema (first 10 for demo, full list has 58+)
gdelt_cols <- c(
  "GLOBALEVENTID", "Day", "MonthYear", "Year", "FractionDate", 
  "Actor1Code", "Actor1Name", "Actor1CountryCode", "Actor1KnownGroupCode", 
  "Actor1EthnicCode", "Actor1Religion1Code", "Actor1Religion2Code", 
  "Actor1Type1Code", "Actor1Type2Code", "Actor1Type3Code",
  "Actor2Code", "Actor2Name", "Actor2CountryCode", "Actor2KnownGroupCode", 
  "Actor2EthnicCode", "Actor2Religion1Code", "Actor2Religion2Code", 
  "Actor2Type1Code", "Actor2Type2Code", "Actor2Type3Code",
  "IsRootEvent", "EventCode", "EventBaseCode", "EventRootCode", 
  "QuadClass", "GoldsteinScale", "NumMentions", "NumSources", 
  "NumArticles", "AvgTone", "Actor1Geo_Type", "Actor1Geo_FullName", 
  "Actor1Geo_CountryCode", "Actor1Geo_ADM1Code", "Actor1Geo_ADM2Code",
  "Actor1Geo_Lat", "Actor1Geo_Long", "Actor1Geo_FeatureID", "Actor2Geo_Type", 
  "Actor2Geo_FullName", "Actor2Geo_CountryCode", "Actor2Geo_ADM1Code",
  "Actor2Geo_ADM2Code", "Actor2Geo_Lat", "Actor2Geo_Long", "Actor2Geo_FeatureID", 
  "ActionGeo_Type", "ActionGeo_FullName", "ActionGeo_CountryCode", 
  "ActionGeo_ADM1Code", "Action2Geo_ADM2Code", "ActionGeo_Lat", "ActionGeo_Long", 
  "ActionGeo_FeatureID", "DATEADDED", "SOURCEURL"
)

# Read the file with tab delimiter and custom column names
gdelt_df <- read_delim(csv_file, delim = "\t", col_names = gdelt_cols, show_col_types = FALSE)

# Preview first rows
head(gdelt_df)
```

**Load GDELT Reference Table Data**

Define reference table data file sources
```{r}
cameo_country_code_txt <- "https://www.gdeltproject.org/data/lookups/CAMEO.country.txt"
fips_country_code_txt <- 'https://www.gdeltproject.org/data/lookups/FIPS.country.txt'
cameo_code_type_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.type.txt'
cameo_known_code_groups_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.knowngroup.txt'
cameo_ethnic_code_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.ethnic.txt'
cameo_religion_code_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.religion.txt'
cameo_event_code_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.eventcodes.txt'
cameo_gold_stein_scale_txt <- 'https://www.gdeltproject.org/data/lookups/CAMEO.goldsteinscale.txt'
adm1_codes_txt <- 'http://efele.net/maps/fips-10/data/fips-414.txt'
adm2_codes_txt <- 'https://download.geonames.org/export/dump/admin2Codes.txt'
```

Actor Data Reference Tables
```{r actor-ref}
# Actor Country Code column reference values
cameo_country_ref <- read.csv(
  cameo_country_code_txt, sep="\t", header=T,
  col.names = c('Country', 'Country_Desc'))

# Actor Known Code Group column reference values
cameo_known_code_groups_ref <- read.csv(
  cameo_known_code_groups_txt, sep="\t", header=T,
  col.names = c('KnownGroup', 'KnownGroup_Desc'))

# Actor Ethnic Code column reference values
cameo_ethnic_ref <- read.csv(cameo_ethnic_code_txt, sep="\t", header=T,
                            col.names = c('Ethnic', 'Ethnic_Desc'))

# Actor Religion Code (1 and 2) column reference values
cameo_religion_ref <- read.csv(cameo_religion_code_txt, sep="\t", header=T,
                              col.names = c('Religion', 'Religion_Desc'))

# Actor Type Code (1, 2, and 3) column reference values
cameo_type_ref <- read.csv(cameo_code_type_txt, sep="\t", header=T,
                                col.names = c('TypeCode', 'Type'))
```

Event Data Reference Tables
```{r}
# Event Code (including base and root event codes) column reference values
cameo_event_ref <- read.csv(cameo_event_code_txt, sep="\t", header=T ,
                           colClasses = "character",
                                col.names = c('EventCode', 'Event'))

# Event Gold Stein Scale column reference values
cameo_gold_stein_ref <- read.csv(cameo_gold_stein_scale_txt, sep="\t", header=T,
                                col.names = c('EventCode', 'GoldSteinScale'))

# Event quad class (primary classification) column reference values
quadclass_ref <- tribble(
  ~"QuadClass", ~"Quad",
  1, 'Verbal Cooperation',
  2, 'Material Cooperation',
  3, 'Verbal Conflict',
  4,' Material Conflict'
)
```

Event Geography Data Reference Tables
```{r event-ref}
# Event geographic resolution column reference values
geo_type_ref <- tribble(
  ~"Geo_Type", ~"Geo_Type_Desc",
  1, 'COUNTRY',
  2, 'USSTATE',
  3, 'USCITY',
  4,' WORLDCITY',
  5, 'WORLDSTATE'
)

# FIPS country column reference values
fips_country_ref <- read.csv(fips_country_code_txt, sep="\t", header=F,
                            col.names = c('GEO_Country', 'GEO_Country_Desc'))

# GEO ADM1 code column reference values
# create U.S. adm1 codes using 'US' followed by the state abbreviation
us_fips_adm1_ref <- tibble(
  'Country' = 'US',
  'division' = 'state',
  fips_adm1 = state.name,
  state_abb = state.abb
  ) |>
  mutate(fips_adm1_code = paste0(Country, state_abb)) |>
  select(fips_adm1_code, division, fips_adm1)

# read in global adm1 codes and join to custom U.S. codes
fips_adm1_ref <- read.csv(
  adm1_codes_txt, sep="_", header=F, na.strings = "",
  col.names = c('fips_adm1_code',
                'first fips version',
                'last fips version',
                'division',
                'designation_2',
                'designation_3',
                'designation4',
                'fips_adm1',
                'convential name',
                'former name')) |>
  select('fips_adm1_code', 'division', 'fips_adm1') |>
  rbind(us_fips_adm1_ref)
```


## Tidy & Clean Data

### 1. Google BigQuery – Top 25 Search Terms

**Google Search Data Cleaning**

  - Transforms `dma_id` into a string dtype
  - Fills Null top_score values with zero
  - Unnests `rising_terms_gains` column to create a long-format dataframe where
    each rising term and associated percent gain is a separate row.
  - Splits `rising_terms_gains` into `rising_term` and `rising_percent_gain` to
    make each value its own column

```{r google-data-cleaning}
long_google_search_df <-  unnest_google_search_df |>
  mutate(
    score = replace_na(score, 0)
    ) |> 
  mutate(search_id = row_number())

head(long_google_search_df)
```

**Google Search Data Tidying**

Subset refresh date into its own table for reference. This is the same for all
rows so only one value is needed to store,
```{r google-refresh-date-ref-table}
refresh_date_ref <- long_google_search_df |>
  select(refresh_date) |>
  distinct()

head(refresh_date_ref)
```

Create Designated Market Area (DMA) reference table.
```{r google-dma-ref-table}
dma_ref_df <- long_google_search_df |>
  select(dma_id, dma_name) |>
  distinct()

head(dma_ref_df)
```

Create final Google top term dateframe
```{r google-top-term-table}
top_term_df <- long_google_search_df |>
  filter(term_type == "top_terms")

head(top_term_df)
```

Create final Google rising term dataframe
```{r google-rising-term-table}
rising_term_df <- long_google_search_df |>
  filter(term_type == "rising_terms")

head(rising_term_df)
```


--
### 2. GDELT 2.0: Our Global World in Realtime

**GDELT Data Cleaning**

  - Transforms `Day` into date data type
  - Transforms `MonthYear` into datetime string format
  - Transforms `GLOBALEVENTID` into strings
  - Transforms all categorical columns into factor data type

```{r gdelt-data-cleaning}
gdelt_df_cleaned <- gdelt_df |>
  mutate_at(vars(contains('Code')), ~as.factor(.)) |>
  mutate_at(vars(contains('Geo_Type')), ~as.factor(.)) |>
  mutate(
    GLOBALEVENTID = as.character(GLOBALEVENTID),
    Day = as.Date(as.character(Day), '%Y%m%d'),
    MonthYear = format(Day, '%Y-%m'),
    QuadClass = as.factor(QuadClass),
  ) |> 
  filter(Day >= today() - days(14) & Day <= today())
```

**GDELT Data Tidying**

No tidying was needed on this dataframe.


---
## Matching the Google Big Query Data with the GDELT Data
```{r join-tbls}
gsearch_gdelt <- 
  left_join(long_google_search_df, 
            gdelt_df_cleaned,
            by=c('refresh_date'='Day'))
```

## Exploratory Data Analysis

*Stop Words* 
Stop words are words that should be excluded in text analysis, such as "the",
"a", "and", etc. Within the 'tidytext' package, stop words can be removed in 
multiple languages. In this analysis, stop words were removed in English and 
Spanish.

```{r stop-words}
# getting stop words for english and spanish 
multi_lang_stopwords <- bind_rows(
  get_stopwords(language = "en"),
  get_stopwords(language = "es")
)
```

*Tokenization*
In text mining, tokenization refers to breaking down text into smaller units.
The google search data was already broken in to short text, therefore, when 
tokenizing the text, the unit of our text becomes a word. 

```{r tokenization}
tokenized_gsearch_gdelt <- gsearch_gdelt |> 
  unnest_tokens(output = word, input = term) |> 
  anti_join(multi_lang_stopwords, by = "word") |> 
  relocate(search_id, word)
```

*Frequency Analysis* 
After tokenizing the text, a freqency analysis is conducted to better understand
which searches appear most frequently, at the word level, and at the Google 
search level. 
```{r count-most-freq-words}
most_freq_tokens <- tokenized_gsearch_gdelt |> 
  count(word, sort = TRUE) |> 
  slice_max(order_by = n, n = 25) 

most_freq_searches <- gsearch_gdelt |> 
  count(term, sort = TRUE) |> 
  slice_max(order_by = n, n = 25)
```

```{r 25-most-freq-tokens-plot}
ggplot(most_freq_tokens, aes(x=reorder(word, -n), y=n))+
  geom_bar(stat="identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
   labs(y = "Count",
       x = "Tokenized Terms (Words)",
       title = "Word Frequency in Google Searches",
       subtitle = "Searches are limited to NYC and LA and are taken from the past two weeks")+
  guides(fill=FALSE)
```

```{r}
ggplot(most_freq_searches, aes(x=reorder(term, -n), y=n))+
  geom_bar(stat="identity")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
   labs(y = "Count",
       x = "Google Searches",
       title = "Google Searches Frequency",
       subtitle = "Searches are limited to NYC and LA and are taken from the past two weeks")+
  guides(fill=FALSE)
```

## Frequency Plots

```{r plot-most-freq-tokens}
tokenized_gsearch_gdelt |> 
  group_by(refresh_date) |> 
  count(word, sort = TRUE) |> 
  top_n(5) |> 
  ungroup() |> 
  mutate(word = reorder_within(word, n, refresh_date),
         date = factor(refresh_date, levels=sort(unique(tokenized_gsearch_gdelt$refresh_date),decreasing=TRUE),ordered = TRUE))|>
  ggplot(aes(x = word, y = n, fill = word)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ date, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()+
  labs(y = "Count",
       x = "Unique words",
       title = "Most frequently googled words in past two weeks",
       subtitle = "Stop words removed from the list")
```

```{r plot-most-freq-top-searches}
gsearch_gdelt |> 
  filter(term_type == "top_terms") |> 
  group_by(refresh_date) |> 
  count(term, sort = TRUE) |> 
  top_n(3) |> 
  ungroup() |> 
  mutate(word = reorder_within(term, n, refresh_date),
         date = factor(refresh_date, levels=sort(unique(gsearch_gdelt$refresh_date),decreasing=TRUE),ordered = TRUE))|>
  ggplot(aes(x = term, y = n, fill = word)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ date, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()+
  labs(y = "Count",
       x = "Top Searched Terms",
       title = "Top Google Searches in Past Two Weeks in LA and NYC",
       subtitle = "")
```

```{r plot-most-freq-rising-searches}
gsearch_gdelt |> 
  filter(term_type == "rising_terms") |> 
  group_by(refresh_date) |> 
  count(term, sort = TRUE) |> 
  top_n(3) |> 
  ungroup() |> 
  mutate(word = reorder_within(term, n, refresh_date),
         date = factor(refresh_date, levels=sort(unique(gsearch_gdelt$refresh_date),decreasing=TRUE),ordered = TRUE))|>
  ggplot(aes(x = term, y = n, fill = word)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ date, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()+
  labs(y = "Count",
       x = "Top Searched Terms",
       title = "Rising Google Searches in Past Two Weeks in LA and NYC",
       subtitle = "")
```

*Word Stemming* 
Another technique that can be used in text mining is word stemming. When word 
stemming, it takes the text, or token, and reduces the term to the root form 
(also known as the stem). This can help normalize the terms, so that terms like
play, playing, played, plays, would all be stemmed to play. 

```{r word-stem}
wrd_stem <- tokenized_gsearch_gdelt %>% 
  mutate(word = SnowballC::wordStem(word))
```

*Document Term Matrix*
A document term matrix represents the frequency of terms that appear in the 
collect of text. 
```{r dtm-top-term}
# creating document term matrix 
dtm_tokenized_terms <- wrd_stem |> 
  count(refresh_date, word) |> 
  cast_dtm(document = refresh_date, term = word, value = n)

# Convert to reg matrix
mat <- as.matrix(dtm_tokenized_terms)
# Convert to sparse Matrix (dgCMatrix) so can use in lapply below
dtm_mat <- Matrix(mat, sparse = TRUE)
```

*Latent Dirichlet Allocation (LDA)* 
LDA is a topic modeling method that can be used to identify central topics and
distributions across a corpus.

**Corpus**: a corpus is a collection of text that is used in natural language 
processing.

In LDA, the value k is defined as the number of topics that appear in the corpus.
To determine k, some additional analysis should be completed. 

```{r determine-k}
# developing list of potential k values to fit in LDA model
k_list <- seq(2, 20, by = 1)

# testing fit for k values in LDA models
fit_test_k_models <- lapply(k_list, function(k) {
  FitLdaModel(dtm = dtm_mat, #using converted matrix
              k = k,
              iterations = 200,
              burnin = 175,
              alpha = 0.1,
              beta = 0.05,
              optimize_alpha = TRUE,
              calc_likelihood = TRUE,
              calc_coherence = TRUE,
              calc_r2 = TRUE)
})
names(fit_test_k_models) <- paste0("k_", k_list)

# Extracting final log-likelihood value
log_likelihood_list <- sapply(fit_test_k_models, function(x) tail(x$log_likelihood, 1)[2])
log_likelihood <- as.numeric(log_likelihood_list)

# generating summary of model perf.
model_summary <- data.frame(
  k = k_list,
  coherence = sapply(fit_test_k_models, function(x) mean(x$coherence)), # topic coherence
  r2 = sapply(fit_test_k_models, function(x) x$r2), #residuals squared - how different are the expected versus obs.
  log_likelihood = log_likelihood # probability that the model generated obs. data
)

ggplot(model_summary, aes(x = k)) +
  geom_line(aes(y = coherence), color = "blue") +
  geom_line(aes(y = r2 * max(coherence)), color = "green", linetype = "dashed") +
  geom_line(aes(y = log_likelihood / max(abs(log_likelihood)) * max(coherence)), 
            color = "red", linetype = "dotted") +
  ylab("Normalized Metric (for comparison)") +
  ggtitle("LDA Tuning: Coherence, R2, and Log-Likelihood")
```

```{r lda-top-term-k15}
# this is where we use a LDA model - the number of topics is defined as K. 
lda_model_k15 <- LDA(dtm_tokenized_terms, k = 15, control = list(seed = 1234))
topics_lda_model_k15 <- tidy(lda_model, matrix = "beta")

# Top terms per topic
top_words_k15 <- topics_lda_model |>   
  group_by(topic) |> 
  top_n(10, beta) |> 
  ungroup()
```

```{r plot-k15}
top_terms_lda_k15 <- topics_lda_model |> 
  group_by(topic) |> 
  slice_max(beta, n = 3, with_ties = FALSE) |> 
  ungroup() |> 
  arrange(topic, -beta)

top_terms_lda_k15 |> 
  mutate(term = reorder_within(term, beta, topic)) |> 
  group_by(topic, term) |>     
  arrange(desc(beta)) |>   
  ungroup() |> 
  ggplot(aes(beta, term, fill = as.factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  labs(title = "Top 5 terms in each LDA topic",
       x = expression(beta), y = NULL) +
  facet_wrap(~ topic, ncol = 4, scales = "free")
```
```{r lda-top-term-k11}
# this is where we use a LDA model - the number of topics is defined as K. 
lda_model_k15 <- LDA(dtm_tokenized_terms, k = 11, control = list(seed = 1234))
topics_lda_model_k15 <- tidy(lda_model, matrix = "beta")

# Top terms per topic
top_words_k11 <- topics_lda_model |>   
  group_by(topic) |> 
  top_n(10, beta) |> 
  ungroup()
```

```{r plot-k11}
top_terms_lda_k11 <- topics_lda_model |> 
  group_by(topic) |> 
  slice_max(beta, n = 3, with_ties = FALSE) |> 
  ungroup() |> 
  arrange(topic, -beta)

top_terms_lda_k11 |> 
  mutate(term = reorder_within(term, beta, topic)) |> 
  group_by(topic, term) |>     
  arrange(desc(beta)) |>   
  ungroup() |> 
  ggplot(aes(beta, term, fill = as.factor(topic))) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  labs(title = "Top 5 terms in each LDA topic",
       x = expression(beta), y = NULL) +
  facet_wrap(~ topic, ncol = 4, scales = "free")
```
## STM Modeling
```{r stm-model}
# temp <- textProcessor(gsearch_gdelt$term, 
#                       metadata = NULL,  
#                       lowercase=TRUE, 
#                       removestopwords=TRUE, 
#                       removenumbers=TRUE,  
#                       removepunctuation=TRUE, 
#                       wordLengths=c(3,Inf),
#                       stem=TRUE,
#                       onlycharacter= FALSE, 
#                       striphtml=TRUE, 
#                       customstopwords=NULL)
# 
# docs <- temp$documents 
# vocab <- temp$vocab 
# 
# tds_stm <- stm(documents=docs, 
#                vocab=vocab, 
#                data = NULL,
#                K=11,
#                max.em.its=25,
#                verbose = FALSE,
#                gamma.prior='L1')
```

## Analyzing with GDELT Data




