---
title: "Project 3: Project Proposal"
author: "Group Members: Alina Vikhnevich, Olivia Azevedo, Alyssa Gurkas, Musrat Jahan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringi)
library(DiagrammeR)
library(dplyr)
library(tidyr)
library(stringr)
```

# Proposal

## Group Members

-   Alina Vikhnevich
-   Olivia Azevedo
-   Alyssa Gurkas
-   Musrat Jahan

## Collaboration Tools

-   ***Group Chat*** **(Slack)** – communication between team members
-   ***GitHub*** – code sharing
-   ***Microsoft Office 365*** – project planning, documentation, and
    presentation development
-   ***R Studio*** – tidying, transforming, and analyzing data

## Data Sources

-   [Computer and Mathematical Occupations – Profile Data](https://www.bls.gov/oes/current/oes150000.htm) - This data
    source provides detailed information on computer and mathematical
    occupations. In this project, we may use the Bureau of Labor
    Statistics to identify occupational codes related to data science,
    and relevant information such as income.
-   [GitHub Archive](https://www.gharchive.org/) - This data source
    includes information about GitHub activity and may be used to track
    which languages developers are using, key terms in projects, and
    committed changes.
-   [Projections Central](https://projectionscentral.org/directdownloads) - This data
    source includes projections of industry and occupational employment
    by state and the US. This could be used to explore the projected
    outcome for certain occupations related to data science such as data
    scientists, analysts, data engineers, and data architects.
-   [O\*Net Database](https://www.onetcenter.org/database.html#all-files) – The
    O\*NET database outlines various information that describe work and
    worker characteristics, including skill requirements for many
    occupations. This data source may be used to explore skillsets,
    applications, and programming languages used in occupations related
    to data science.

## Read Data

```{r load-data}
# load the O*NET Skills Dataset, data dictionary available here: 
# https://www.onetcenter.org/dictionary/25.1/excel/knowledge.html
skills = read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/main/Project%203/Skills.csv")

# load the OES Data [insert info on where data came from]
oes = read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/main/Project%203/OES_Report_Industries_with_DS_Employment.csv",
                  skip = 6, nrows = 284,
                  header = FALSE)

# load the United Nations Standard Products and Services Code, exported from here: https://www.ungm.org/public/unspsc
soc_ref = read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/main/Project%203/UNSPSC%20Reference.csv")
```

```{r col-names-soc-ref}
colnames(soc_ref) = c(
                        'commodity_code', 
                        'commodity_title', 
                        'class_code', 
                        'class_title', 
                        'family_code', 
                        'family_title', 
                        'segment_code', 
                        'segment_title')

soc_ref = soc_ref |>
  mutate_all(as.character)
```

```{r col-names-skills}
colnames(skills) = c('class_code', 'class_title', 'element_id', 'element_name', 'scale_id', 'scale_name',
                     'value', 'n', 'se', 'lower_ci', 'upper_ci', 'recommend_suppress', 
                      'not_relevant', 'date', 'domain_source')

skills_df = skills |>
  mutate(class_code = str_replace(str_replace(class_code, '\\.', ''), '-', ''))
```

```{r col-names-oes}
colnames(oes) = c('industry', 'employment', 'employment_prct_relative_se',
                             'hr_wage_mean', 'annual_wage_mean', 'wage_prct_relative_se','hr_wage_10_prcentile',
                              'hr_wage_25_prcentile', 'hr_wage_median', 'hr_wage_75_prcentile',
                              'hr_wage_90_prcentile', 'annual_wage_10_prcentile',
                              'annual_wage_25_prcentile', 'annual_wage_median', 'annual_wage_75_prcentile',
                              'annual_wage_90_prcentile')


oes_df = oes |>
  mutate(soc = str_match(industry, "\\s*(\\d{2}-\\d{2,}[A-Z]*\\d*)")[,2]) |>
  mutate(soc = stri_pad_right(str_replace(soc, '-', ''), 8, 0)) |>
  mutate(sector_soc_1 =  str_match(industry, "^Sector(s)*\\s{1}(.*)\\s-.*")[,3]) |>
  mutate(sector_soc_2 = str_match(industry, "\\s*(\\d{2}---\\d{2})")[,2]) |>
  mutate(sector_soc = ifelse(is.na(sector_soc_1), sector_soc_2, sector_soc_1)) |>
  mutate(sector_soc = str_replace(str_replace(str_replace(sector_soc, '---', ', '), 'and', ', '), ', ,', ',')) |>
  mutate(segment_soc = ifelse(is.na(sector_soc), sector_soc, stri_pad_right(sector_soc, 8, 0))) |>
  mutate(industry = gsub( " \\(.*$", "", industry))
```

## Industry tidy normalized tables

```{r tbls-industry}
family_hr_wage_df = 
  oes_df |>
  dplyr::filter(is.na(segment_soc)) |>
  select(soc, starts_with('hr')) |>
  pivot_longer(
    cols = starts_with('hr'),
    names_to = 'hourly_wage_stat',
    names_prefix = 'hr_wage_',
    values_to = 'dollars'
  ) |>
  mutate(dollars = str_remove(dollars, "^\\$"))  # Removes "$"

family_annual_wage_df = 
  oes_df |>
  dplyr::filter(is.na(segment_soc)) |>
  select(soc, starts_with('annual')) |>
  pivot_longer(
    cols = starts_with('annual'),
    names_to = 'annual_wage_stat',
    names_prefix = 'annual_wage_',
    values_to = 'dollars'
  ) |>
  mutate(dollars = str_remove(dollars, "^\\$"))  # Removes "$"
```

```{r wage-tbls}
segemnt_hr_wage_df = oes_df |>
  drop_na(segment_soc) |>
  select(segment_soc, starts_with('hr')) |>
  pivot_longer(
    cols = starts_with('hr'),
    names_to = 'hr_wage_stat',
    names_prefix = 'hr_wage_',
    values_to = 'dollars'
    ) |>
  mutate(dollars = substr(dollars, 2, length(dollars))) 


segemnt_annual_wage_df = oes_df |>
  drop_na(segment_soc) |>
  select(segment_soc, starts_with('annual')) |>
  pivot_longer(
    cols = starts_with('annual'),
    names_to = 'annual_wage_stat',
    names_prefix = 'annual_wage_',
    values_to = 'dollars'
    ) |>
  mutate(dollars = substr(dollars, 2, length(dollars))) 
```

```{r skills-tbls}
element_ref = 
  skills_df |>
  select(element_id, element_name) |>
  distinct()

im_skills_df = skills_df |>
  dplyr::filter(scale_id == 'IM') |>
  select(class_code, class_title, element_id, value)

level_skills_df = skills_df |>
  dplyr::filter(scale_id == 'LV') |>
  select(class_code, class_title, element_id, value)
```

## Logical Model

The logical model represents the relationships between different
entities in the dataset, specifically focusing on skills, employment
statistics, and job market trends.

**Entities and Attributes:**

-   **Jobs** (`job_id`, `title`, `SOC_code`, `industry`,
    `median_salary`, `projected_growth`)

-   **Skills** (`skill_id`, `name`, `category`, `importance`, `level`)

-   **Job_Skills** (Bridge Table: `job_id`, `skill_id`)

-   **Industries** (`industry_id`, `name`, `description`)

-   **Employment_Statistics** (`job_id`, `industry_id`,
    `hourly_median_wage`, `annual_median_wage`, `employment_count`)
    
```{r logical-model}   
grViz("
digraph Logical_Model {

  graph [layout = dot, rankdir = TB]

  # Define node styles
  node [shape = rectangle, style = filled, fillcolor = lightblue]

  # Data Sources
  DataSources [label = 'Raw Data Sources', shape = parallelogram, fillcolor = lightgray]
  ONET [label = 'O*NET Database \\n(Skills & Job Data)']
  BLS [label = 'BLS Employment Statistics \\n(Job Salaries)']
  SkillsCSV [label = 'Skills.csv \\n(Job-Skill Mapping)']
  IndustryTrends [label = 'Industry Trends Data']

  # Processing
  Processing [label = 'Data Processing & Normalization', shape = ellipse, fillcolor = lightyellow]
  Cleaning [label = 'Data Cleaning & \\n Transformation']
  Normalization [label = 'Normalization of \\n Job & Skill Data']

  # Structured Data Tables (Fixed Black Block)
  StructuredData [label = 'Final Structured Data', shape = parallelogram, fillcolor = lightgray, style = filled]
  Jobs [label = 'Jobs Table']
  Skills [label = 'Skills Table']
  Employment [label = 'Employment Statistics Table']
  JobSkills [label = 'Job-Skills Mapping Table']

  # Relationships
  DataSources -> {ONET BLS SkillsCSV IndustryTrends}
  {ONET BLS SkillsCSV IndustryTrends} -> Processing
  Processing -> {Cleaning Normalization}
  {Cleaning Normalization} -> StructuredData
  StructuredData -> {Jobs Skills Employment JobSkills}
}
")
```

## Entity Relationship Diagram

Entity Relationship Diagrams specify the nature of the relationship
between tables in a database.

```{r er-diagram}
grViz("
digraph ER_Diagram {
  
  graph [layout = dot, rankdir = LR]

  # Define node styles
  node [shape = rectangle, style = filled, fillcolor = lightblue]

  # Entities
  Jobs [label = 'Jobs\\n(job_id, title, SOC_code, industry, median_salary, projected_growth)']
  Skills [label = 'Skills\\n(skill_id, name, category, importance, level)']
  Job_Skills [label = 'Job_Skills\\n(job_id, skill_id)']
  Industries [label = 'Industries\\n(industry_id, name, description)']
  Employment_Statistics [label = 'Employment_Statistics\\n(job_id, industry_id, hourly_median_wage, annual_median_wage, employment_count)']

  # Relationships
  Jobs -> Job_Skills [label = 'Many-to-Many']
  Skills -> Job_Skills [label = 'Many-to-Many']
  Jobs -> Industries [label = 'Many-to-One']
  Jobs -> Employment_Statistics [label = 'One-to-Many']
  Industries -> Employment_Statistics [label = 'One-to-Many']
}
")
```
