---
title: "Project 3: Project Proposal"
author: "Group Members: Alina Vikhnevich, Olivia Azevedo, Alyssa Gurkas, Musrat Jahan"
date: "`r Sys.Date()`"
format:
  revealjs:
    theme: cerulean
    toc: true
    toc-depth: 2
    slide-level: 2
    transition: fade
execute:
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This project explores what are the most valuable data science skills. To answer this, the following methodology was followed:

- **Data Collection** - Import data from various sources such as Bureau of Labor Statistics, Projections Central, and O*Net.
- **Data Normalization** - Clean and normalize the data using various processing techniques.
- **Export to Database** - Store the processed data.
- **Data Analysis** - Conduct analysis on the structured data.
- **Summary of Findings** - Summarize key insights.

### Research Questions

1. Which skills are considered the most important in the data science field?
2. What is the relationship between skill importance and job compensation? How does this vary across industries?
3. What is the relationship between projected employment and the importance of job skills?

## Data Sources

- [Industry Profile for Data Scientists](https://www.bls.gov/oes/current/oes150000.htm)
- [Projections Central](https://projectionscentral.org/directdownloads)
- [O\*Net Database](https://www.onetcenter.org/database.html#all-files)

## Logical Model

## Entity Relationship Diagram
*Data Normalization* 
To reduce redundancy and improve data integrity, the data in this project was 
normalized. This helps to ensure that data is stored efficiently, avoiding 
duplication and inconsistencies, and to have a better-managed database. To 
normalize the datasets, five core tables were developed, and five reference 
tables were developed. 

## Unique Identifiers
This project pulls data from three sources and matches on two IDs: the Standard 
Occupational Classification (SOC) and the North American Industry Classification 
System (NAICS) codes. The SOC codes work as a tree structure. The more digits 
added to the 6-digit SOC code (or 6 digit with 2 decimals O*NET soc code), the 
more in depth the occupation description becomes.  

## Load Data

```{r load-data}
# insert connection to relational database
```



## Tidy Data
To tidy the data the team: 
1. Renamed columns for more intuitive names 
2. Developed reference tables to store distinct categorical values (e.g., skill categories).  
3. Removed redundant columns from the core data tables, retaining only relevant fields. 
4. Transformed certain datasets into a long format for better usability, such as restructuring wage statistics into separate columns for duration, type, and value.


## Data Analysis

```{r analyze-data}
# insert code analyzing data 

```

## Findings

```{r display-findings}
# insert visualizations
```

## Conclusion

Summarize the key insights and implications of the study.

---