---
title: "Project 3: Project Proposal"
author: "Group Members: Alina Vikhnevich, Olivia Azevedo, Alyssa Gurkas, Musrat Jahan"
date: "`r Sys.Date()`"
format:
  revealjs:
    theme: cerulean
    toc: true
    toc-depth: 2
    slide-level: 2
    transition: fade
execute:
  echo: true
---

```{r load-libraries}
library(tidyverse)
library(stringi)
library(DiagrammeR)
library(dplyr)
library(tidyr)
library(stringr)
library(kableExtra)
library(jsonlite)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
tech_skills <- read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/tidy_normalize_data_oa/Project%203/Data/Technology%20Skills.csv")

# load additional occupation skills data
ep_project_skills <- read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/tidy_normalize_data_oa/Project%203/Data/public-skills-data.csv", skip = 1)

# load occupation project data
soc_industry_project <- read.csv("https://raw.githubusercontent.com/AlinaVikhnevich/data_607/refs/heads/tidy_normalize_data_oa/Project%203/Data/National_Employment_Matrix_for_2023_and_projected_2033.csv")

# defining
data_science_soc = '15-2051'

# updated column names
colnames(tech_skills) <- c(
  'onet_soc', 'onet_soc_title', 'tech_skill_example',
  'commodity_code', 'commodity_title', 'hot_technology',
  'in_demand')

# rendering table 
knit_table(head(tech_skills), 'View Raw Tech Skills Data')

```


## Introduction
This project explores what are the most valuable data science skills. To answer 
this, the following methodology was followed:

- **Data Collection** - Import data from various sources such as Bureau of Labor 
Statistics, Projections Central, and O*Net.
- **Data Normalization** - Clean and normalize the data using various processing 
                           techniques.
- **Export to Database** - Store the processed data.
- **Data Analysis** - Conduct analysis on the structured data.
- **Summary of Findings** - Summarize key insights.

### Research Questions

1.  Which skills are considered the most important in the data science
    field?
2.  What is relationship between projected employment and the importance
    of job skills?
3.  How does the presence of 'hot technologies' relate to skill
    importance in data science roles?
4.  What types of technical skills (based on commodity categories) are
    most commonly marked as in-demand or hot?
5.  What is the distribution of skill importance across different skill
    categories (e.g., cognitive, interpersonal)?

## Data Sources

-   [Industry Profile for Data Scientists](https://www.bls.gov/oes/current/oes150000.htm) - This
    data source provides detailed information on the data science
    occupation. This project uses the Bureau of Labor Occupational
    Employment and Wage (OES) statistics to identify industries with
    employment in Data Scientists and relevant information such as
    income.
-   [Projections Central](https://projectionscentral.org/directdownloads) - This data
    source includes projections of industry and occupational employment
    by state and the US. This data was used to explore the relationship
    of skills and projected outcomes for certain occupations related to
    data science such as data scientists, analysts, data engineers, and
    data architects.
-   [O\*Net Database](https://www.onetcenter.org/database.html#all-files) – The
    O\*NET database outlines various information that describe work and
    worker characteristics, including skill requirements for specific
    occupations. This data source was used to explore various skill sets
    for occupations related to data science.

## Logical Model

```{r logical-model}   
grViz("
digraph Logical_Model {

  graph [layout = dot, rankdir = TB]

  # Define node styles
  node [shape = rectangle, style = filled, fillcolor = lightblue]

  # Data Sources
  DataSources [label = 'Raw Data Sources', shape = parallelogram, fillcolor = lightgray]
  ONET [label = 'O*NET Database \\n(Skills & Job Data)']
  BLS [label = 'Bureau of Labor Statistics \\n(Employment Statistics for Data Scientists)']
  Projections [label = 'Projections \\n(Long-term projections for employment)']

  # Processing
  Processing [label = 'Data Processing & Normalization', shape = ellipse, fillcolor = lightyellow]
  Cleaning [label = 'Data Cleaning & \\n Transformation']
  Normalization [label = 'Normalization of \\n Job & Skill Data']

  # Structured Data Tables (Fixed Black Block)
  StructuredData [label = 'Final Structured Data', shape = parallelogram, fillcolor = lightgray, style = filled]
  Core [label = 'Core Tables']
  Link [label = 'Link Tables']
  Ref [label = 'Reference Tables']
  

  # Relationships
  DataSources -> {ONET BLS Projections}
  {ONET BLS Projections} -> Processing
  Processing -> {Cleaning Normalization}
  {Cleaning Normalization} -> StructuredData
  StructuredData -> {Core Link Ref}
}
")
```


## Entity Relationship Diagram
```{r erd}
grViz("
digraph ER_Diagram {
  graph [layout = dot, rankdir = LR, fontname = 'Helvetica']
  node [shape = rectangle, style = filled, fontname = 'Helvetica']

  // Core Tables Container
  subgraph cluster_core {
    label = 'Core Tables';
    style = 'rounded,filled';
    color = gray;
    fillcolor = white;
    // Nodes in Core cluster
    EpSkills [label = 'ep_skills\\n(soc,\\nep_skills_category_id,\\nep_skills_score)', fillcolor = '#4F7942', fontcolor = white]
    ONETSkills [label = 'onet_skills\\n(soc,\\nonet_element_id,\\nonet_rating_value)', fillcolor = '#4F7942', fontcolor = white]
    TechSkills [label = 'tech_skills\\n(soc,\\ncommodity_code,\\nhot_technology,\\nin_demand)', fillcolor = '#4F7942', fontcolor = white]
    SocIndustryProject [label = 'soc_industry_project\\n(industry_code,\\nsoc,\\nemployment,\\nprct_industry,\\nprct_soc,\\nyear)', fillcolor = '#4F7942', fontcolor = white]
    SocIndustryProjectChange [label = 'soc_industry_project_change\\n(industry_code,\\nsoc,\\nemployment_2023,\\nprct_soc_2023,\\nemployment_2033,\\nemployment_change_num)', fillcolor = '#4F7942', fontcolor = white]
  }

  // Reference Tables Container
  subgraph cluster_ref {
    label = 'Reference Tables';
    style = 'rounded,filled';
    color = gray;
    fillcolor = white;
    // Nodes in Reference cluster
    CommodityRef [label = 'commodity_ref\\n(commodity_code,\\ncommodity_title)', fillcolor = '#8fc0a9', fontcolor = black]
    SkillsElementRef [label = 'skills_element_ref\\n(onet_element_id,\\nonet_element_name)', fillcolor = '#8fc0a9', fontcolor = black]
    SocRef [label = 'soc_ref\\n(soc_title,\\nsoc,\\nsoc_type)', fillcolor = '#8fc0a9', fontcolor = black]
    SkillsCategoryRef [label = 'skills_category_ref\\n(ep_skills_category_id,\\nep_skills_category)', fillcolor = '#8fc0a9', fontcolor = black]
    IndustryRef [label = 'industry_ref\\n(industry_code,\\nindustry_title,\\nindustry_type)', fillcolor = '#8fc0a9', fontcolor = black]
  }

  // Link Tables Container
  subgraph cluster_link {
    label = 'Link Tables';
    style = 'rounded,filled';
    color = gray;
    fillcolor = white;
    // Nodes in Link cluster
    SocOnetLink [label = 'soc_onet_soc_lnk\\n(soc,\\nonet_soc)', fillcolor = '#faf3dd', fontcolor = black]
    SocIndustryLink [label = 'soc_industry_lnk\\n(soc,\\nindustry_code)', fillcolor = '#faf3dd', fontcolor = black]
  }

  // Ensure horizontal alignment for nodes within each cluster
  { rank = same; EpSkills; ONETSkills; TechSkills; SocIndustryProject; SocIndustryProjectChange; }
  { rank = same; CommodityRef; SkillsElementRef; SocRef; SkillsCategoryRef; IndustryRef; }
  { rank = same; SocOnetLink; SocIndustryLink; }

  // Relationships: All tables with a 'soc' column link to SocRef
  EpSkills -> SocRef [label = 'soc', fontcolor = blue]
  ONETSkills -> SocRef [label = 'soc', fontcolor = blue]
  TechSkills -> SocRef [label = 'soc', fontcolor = blue]
  SocIndustryProject -> SocRef [label = 'soc', fontcolor = blue]
  SocIndustryProjectChange -> SocRef [label = 'soc', fontcolor = blue]
  SocOnetLink -> SocRef [label = 'soc', fontcolor = blue]
  SocIndustryLink -> SocRef [label = 'soc', fontcolor = blue]

  // Other Relationships
  EpSkills -> SkillsCategoryRef [label = 'ep_skills_category_id', fontcolor = blue]
  ONETSkills -> SkillsElementRef [label = 'onet_element_id', fontcolor = blue]
  TechSkills -> CommodityRef [label = 'commodity_code', fontcolor = blue]
  SocIndustryProject -> IndustryRef [label = 'industry_code', fontcolor = blue]
  SocIndustryProjectChange -> IndustryRef [label = 'industry_code', fontcolor = blue]

  // Link Tables relationships
  SocOnetLink -> ONETSkills [label = 'onet_soc', fontcolor = blue]
  SocIndustryLink -> IndustryRef [label = 'industry_code', fontcolor = blue]
}
")
```


## Data Normalization* 
To reduce redundancy and improve data integrity, the data in this project was 
normalized. This helps to ensure that data is stored efficiently, avoiding 
duplication and inconsistencies, and to have a better-managed database. To 
normalize the datasets, five core tables were developed, and five reference 
tables were developed. 

## Unique Identifiers
This project pulls data from three sources and matches on two IDs: the Standard 
Occupational Classification (SOC) and the North American Industry Classification 
System (NAICS) codes. The SOC codes work as a tree structure. The more digits 
added to the 6-digit SOC code (or 6 digit with 2 decimals O*NET soc code), the 
more in depth the occupation description becomes.  

## Load Data
```{r load-data, include=FALSE}
# Core tables
ep_skills_df_clean <- read_csv("Data/ep_skills_df_clean.csv")
onet_skills_df_clean <- read_csv("Data/onet_skills_df_clean.csv")
tech_skills_df_clean <- read_csv("Data/tech_skills_df_clean.csv")
soc_industry_project_df_clean <- read_csv("soc_industry_project_df_clean.csv")
soc_industry_project_change_df_clean <- read_csv("soc_industry_project_change_df_clean.csv")

# Link tables
soc_onet_soc_lnk <- read_csv("Data/soc_onet_soc_lnk.csv")
soc_industry_lnk <- read_csv("Data/soc_industry_lnk.csv")

# Reference tables
commodity_ref <- read_csv("Data/commodity_ref.csv")
skills_element_ref <- read_csv("Data/skills_element_ref.csv")
soc_ref <- read_csv("soc_ref.csv")
skills_category_ref <- read_csv("Data/skills_category_ref.csv")
industry_ref <- read_csv("Data/industry_ref.csv")
```

```{r database-connect, include=FALSE}
# Securely fetching the password (ensure your environment variable is set)
password <- Sys.getenv("MYSQL_PASSWORD")

# Attempt to connect
conn <- tryCatch({
  dbConnect(
    MySQL(),
    user = "root",
    password = password,
    dbname = "project3_data",
    host= '192.168.1.52',
    port= 3306,
    client.flag = CLIENT_LOCAL_FILES
  )
}, error = function(e) {
  message("Error: ", e$message)
  return(NULL)
})
```

```{r database-load, scroll: true}
# Proceed only if connection succeeded
if (!is.null(conn)) {
  print("Database connection successful.")
  
  # Load Core tables
  ep_skills_df_clean <- dbReadTable(conn, "ep_skills")
  onet_skills_df_clean <- dbReadTable(conn, "onet_skills")
  tech_skills_df_clean <- dbReadTable(conn, "tech_skills")
  soc_industry_project_df_clean <- dbReadTable(conn, "soc_industry_project")
  soc_industry_project_change_df_clean <- dbReadTable(conn, "soc_industry_project_change")
  
  # Load Link tables
  soc_onet_soc_lnk <- dbReadTable(conn, "soc_onet_soc_lnk")
  soc_industry_lnk <- dbReadTable(conn, "soc_industry_lnk")
  
  # Load Reference tables
  commodity_ref <- dbReadTable(conn, "commodity_ref")
  skills_element_ref <- dbReadTable(conn, "skills_element_ref")
  soc_ref <- dbReadTable(conn, "soc_ref")
  skills_category_ref <- dbReadTable(conn, "skills_category_ref")
  industry_ref <- dbReadTable(conn, "industry_ref")
  
  # Optionally, print a sample from one table for verification
  print(head(ep_skills_df_clean))
  
  # Disconnect when done
  dbDisconnect(conn)
  
} else {
  stop("Database connection failed. Check credentials and try again.")
}
```

### Tidy Data

To tidy and normalize the data the team performed the following

1.  Renamed columns to allow for more intuitive names as well as ensure
    columns representing the same data values are referenced the same
    across all data frames.

2.  Developed reference tables to store distinct categorical values
    (ex: skill categories) and remove partial dependencies.

3.  Removed redundant columns (such as columns that are represented in
    reference tables) from the core data tables, retaining only relevant
    fields for analysis.

## Data Analysis

```{r analyze-data}
# insert code analyzing data 

```

## Findings

```{r display-findings}
# insert visualizations
```

## Conclusion

Summarize the key insights and implications of the study.

---